---
title: "tidying"
author: "Saneesh"
date: "1/24/2022"
output: pdf_document
---
# packages

```{r gapminder}
# install.packages ('gapminder')
library(gapminder)
library(tidyverse)
```

# shortcuts

alt+- will add <-  
shift+ctrl+c to add # infront of a line  
'----' for a header, so it is easy to navigate through the script  
command +shift + m  
ctrl+alt+i for new code chunk 
# syntax
Plain text  
end a line with two spaces to start a new paragraph.  
*italics* and _italics_  
**bold** and __bold__  
superscript^2^  
~~strikethrough~  
[link](www.rstudio.com) to rstudio

# logical operations

1==1 # equality  
1!=3 #unequal  
13<14 #13 smaller than 14  
14>13 #14 bigger than 13  
12>=0 #12 greater or equal to zero  
12<=3 #12 smaller or equal to zero  

# creating data.frame
family  
```{r echo=TRUE}
name <- c('saneesh', 'sanusha', 'appu', 'kishan')
weight <- c(63,48, 20, NA)
height <- c(164, 150, NA, 75)
family <- data.frame(name, weight, height)
family %>% as_tibble()
```

#### data frame with unequal values 10 and 8
```{r}
library(tidyverse)
data <- data.frame(sex=c(rep('female', 10), rep('male', 8)), 
                   score=c(rnorm(n= 10, mean = 7.56, sd = 1.978), rnorm(n= 8, mean=7.75, sd= 1.631)))

data

data %>% group_by(sex) %>% 
  summarise(score= n()) %>% 
  mutate(freq=score/sum(score)*100)

```

# tibble
```{r}
library(tidyverse)
years <- tribble(
  ~Location, ~Year, ~Month, ~Day,
  "Sydney", 2000, 9, 15,
  "Athens", 2004, 8, 13,
  "Beijing", 2008, 8, 8,
  "London", 2012, 7, 27,
  "Rio de Janeiro", 2016, 8, 5
)
```

# tabyl

[tabyl](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html)

```{r}



library(janitor)
data <- data.frame(HairEyeColor)

data %>% tabyl(Hair, Eye) %>% 
  adorn_percentages('row') %>% 
  adorn_pct_formatting(digits = 2) %>% 
  adorn_ns() %>% 
  knitr::kable()

```



# is.na
```{r}
# identify location of NAs in vector
which(is.na(family))
colSums(is.na(family)) 
```

# replace na
```{r}
mat <- matrix(sample(c(NA, 1:5), 50, replace = TRUE), 5)
df <- as.data.frame(mat)
df %>% replace(is.na(.), 0)%>% view()
```

# clean names

```{r}
# install.packages('janitor')
library(janitor)

id <- (c(1,1,2,2,3,3))
Country <- c('Angola', 'Angola','Botswana', 'Botswana','Zimbabwe','Zimbabwe')
year <- c('2006', '2007', '2008', '2009', '2010', '2006')
bank.ratio <- c(24,25,38,34,42,49)
Reserve.ratio <- c(77,59,64,65,57,86)
broad.money <- c(163,188,317,361,150,288)


bank <- data.frame(id, Country, year, bank.ratio, Reserve.ratio,broad.money)

bank %>% view()
  as_tibble()
  
bank <- bank %>% clean_names() # replaced . with _
  
glimpse(bank)  
```

```{r}
bank <- bank %>% clean_names() # replaced . with _

```


**filter** bank data frame below such that it retains a country if a given id is satisfied e.g. filtering a data frame that has countries with id 1 and 2 only

```{r}
bank %>% 
  filter(id%in% c(1,2)) %>% 
  as_tibble()
```
**summarise** fund available with each countries
```{r echo=TRUE}
bank %>% 
  group_by(country) %>% 
  summarise(fund=sum(broad_money)) %>% 
  as_tibble()

```

# count/ frequency
```{r echo=TRUE}
mtcars %>% 
  count(am) %>% 
  as_tibble()

```
```{r}
mtcars %>% 
  count(gear)
```
# rename column
column: new name= old name

```{r}
iris %>% 
  rename(S.len=Sepal.Length,
         Sp.= Species) %>% head(3)
```

## rename to lower
```{r}
iris %>% 
  rename_with(tolower) %>% head(3)

```

## rename to lower specific columns

```{r}
iris %>% select_at(vars(Species, Petal.Length), tolower) %>% head(3)
```

# add column  
```{r}
library(tibble)
iris %>% add_column(ob_no=1:150)
```

# gapminder 


```{r echo= TRUE}
iris %>% as_tibble()
```


```{r echo= TRUE }
summary(gapminder)
str(gapminder)
```

# recode observation

change name of observation--- mutate (variable=recode (variable, 'old name'='new name')))
```{r}
gapminder %>% 
 mutate(country=recode(country, 'India'='IND' )) %>% 
  filter(country=='IND')

```


# select
```{r echo= TRUE}
gapminder %>% 
  select(year, country, gdpPercap) 
```

```{r}
msleep %>% select(starts_with("sleep"))
```

## do not select

```{r}
iris %>% select(-Sepal.Length, -Species) %>% head(3)
```

or 

```{r}
iris %>% select(-c(Sepal.Length)) %>% head(3)
```
```{r}
iris %>% select(!Sepal.Length) %>% head(3)
```


## `ends_with`
```{r}
iris %>% select(ends_with('length')) %>% head(3)
```
## `starts_with`
```{r}
iris %>% select(starts_with('Sepal')) %>% head(3)
```



# filter 

```{r echo= TRUE }
gapminder %>% 
  select(year, country, lifeExp) %>% 
  filter(country=="Eritrea", year>1950)
```

```{r echo=TRUE}
gapminder %>% filter(country=="Canada") %>% head(3) # from gapminder data filter country Canada and show only 2 observations
```
## except
```{r echo=TRUE }
gapminder %>% filter(country!="Oman") %>% head(3) # from gapminder data filter all the other countries except Oman
```
## omit

```{r}
iris %>% filter(Species!='setosa') %>% glimpse()
```
## omit multiple
```{r}
iris %>% filter(!Species %in% c('setosa', 'versicolor')) %>% glimpse()
```

## filter between
```{r}
iris %>% filter(Petal.Width >=2 & Petal.Width <= 5) %>% glimpse()
```



# pull
```{r}
iris %>% pull(Species) # returns vector values
iris %>% select(Species) # returns a table with one column
iris %>% select(everything()) %>% head(3)


```



## multiple conditions
```{r echo=TRUE}
gapminder %>% 
  filter(country=="Oman" &
           year>1980 &
           year<=2000) %>% head(4)
```

```{r echo=TRUE}
gapminder %>% 
  select(country, year) %>% 
  filter(year>=1980, country=="India"|
           country=="Oman"|
           country=="Canada")
```

```{r}
gapminder %>% filter(country!="Oman") %>% head(3) # from gapminder data filter all the other countires exept Oman

```

## filter multipe  using %in%
```{r}
gapminder %>% filter(country %in% c('Hungary','Iceland', 'Mongolia')) %>% head(3)

target <- c('Hungary','Iceland', 'Mongolia')
gapminder %>% filter(country %in% target) %>% head (3)
```

```{r}
friends <- data.frame(Names=c('Saneesh', 'Appu', 'Shruti', 'Aradhana', 'Arathi', 'James Bond'),
                      age=c(40,9, 25, 25, 25, 50))
# data frame is friends
# columns in friends are Names, Age, Height, etc.
# Colum Name have 'Saneesh', 'Appu', 'Shruti', 'Aradhana', 'Arathi', 'James Bond'
# We want to filter information related to Sanees and James Bond only, so we created a vector with
# these names in it.

target <- c('Appu', 'James Bond') #and then

friends %>% filter(Names %in% target)
# or 
friends %>% filter(Names== 'Appu'| Names== 'James Bond')
# or 
friends %>% filter(Names %in% c('Appu', 'James Bond'))
```


# drop
```{r echo=TRUE}
gapminder %>% 
  select(-year,-pop) %>% 
  head(5)

```


# group by & summarise
```{r echo= TRUE}
gapminder %>% 
  filter(year==2007) %>% 
  group_by(country) %>% 
  summarise(meanLE=mean(lifeExp)) %>% 
  arrange(meanLE,decreasing = TRUE)
```
```{r echo= TRUE}
gapminder %>% 
  group_by(country) %>% 
  summarise(minLE=min(lifeExp)) %>% 
  arrange(minLE,decreasing=FALSE)
```

grouped by continent, then summarise two things, first n=n() number of rows in which each continent are or the size of each group, then the mean of the mean of the lifeExp variable.

```{r echo= TRUE}
gapminder %>% 
  group_by(continent) %>% 
  summarise(n=n(),
            meanLife=mean(lifeExp))
```
```{r echo= TRUE}
gapminder %>% 
  group_by(continent) %>% 
  summarise(PopConti=sum(pop))
```

```{r}
pets <- data.frame(names=c(rep('saneesh', 3), rep('appu', 2), 'sanusha'), 
                   pet=c(rep('dog', 3), rep('cat', 2), 'tiger'), number=c(2,2,5,7,8,1), 
                   size=c(rep('medium', 2), rep('small', 3), 'big'))

pets

library(tidyverse)

pets %>% group_by(pet, size) %>% 
  summarise(totalpet= sum(number))
```

# summarise

```{r}
library(tidyverse)
plot <- c(rep(1,2), rep(2,4), rep(3,3))
bird <- c('a','b', 'a','b', 'c', 'd', 'a', 'b', 'c')
area <- c(rep(10,2), rep(5,4), rep(15,3))

birdlist <- data.frame(plot,bird,area)
birdlist

# summarize the following data frame to a summary table.
# option 1
birdlist %>% 
  group_by(plot) %>% 
  summarise(bird = n(), area = unique(area))

# option 2
birdlist %>%
  count(plot, area, name = "bird")
```


```{r echo= TRUE}
gapminder %>% 
  summarise(mean(lifeExp))
```


```{r echo= TRUE}
gapminder %>%
  summarise(range(lifeExp))
```


```{r echo= TRUE}
gapminder %>% 
  filter(country=="India") %>% 
  group_by(country) %>% 
  summarise(GDPmax=max(gdpPercap),
            GDPmin=min(gdpPercap),
            GDPmean=mean(gdpPercap))
```


# count/summarize
```{r}
library(tidyverse)
plot <- c(rep(1,2), rep(2,4), rep(3,3))
bird <- as.factor(c('a','b', 'a','b', 'c', 'd', 'a', 'b', 'c'))
area <- c(rep(10,2), rep(5,4), rep(15,3))

birdlist <- data.frame(plot,bird,area)
birdlist

#birdlist %>%    group_by(plot, area) %>%    mutate(count(bird))


birdlist %>%
  group_by(plot, area) %>%
  dplyr::summarize(bird = n(), # when summarize doesn't work directly use it (dplyr::)like this
            .groups = "drop") # to summarize of a column with reference to two other variables. 
```

# count sites

```{r}
treatment <- c(rep('ab',2), rep('bgrnf', 8), rep('bgpnf', 4))
site <- c('ab1', 'ab2', 
          rep('bgrnf1', 3), 
          rep('bgrnf2', 2), 
          'bgrnf3', 
          'bgrnf4', 
          'bgrnf5',
          rep('bgpnf1', 2),
          rep('bgpnf2', 2))
data <- data.frame(treatment, site)
library(tidyverse)
# to find the site per each treatment
data %>% group_by(treatment) %>% count(treatment)

```

# case when new column  
```{r}
library(dplyr)
library(stringr)
feedback <- c('good_book', 'good_read', 'good_story', 'good for knowledge')
book <- c('ramayana', 'bible', 'encyclopedia', 'Mbharatha')

df <- data.frame(feedback, book)

df %>%
  mutate(response = case_when(str_starts(feedback, 'good') ~ 'good')) %>% 
  select(book, response) %>% as_tibble()
```




# separate 
text to columns
```{r echo=TRUE}
df <- data.frame(films = c("Spider_man", "James_bond", "Iron_man", "Bat_man"))
df
df1 <- df %>% 
  separate(films, c("a", "b"), sep='([_])')
df1
```
# unite
```{r}
df1 %>% unite("names", a:b, remove=FALSE)
```

# join
```{r}
df2 <- data.frame(id=c(1:4) ,films = c("Spider_man", "James_bond", "Iron_man", "Bat_man"))
df3 <- data.frame(id=c(1:4) ,country= rep("us", 4))

df4 <- left_join(df2, df3, by="id")
df4


```





# across
for multiple variables 
```{r}
library(tidyverse)
srno <- c(1:2)
film <- c("arabica", "robust")
rate <- c("good", "better")
lang_Eng <- c("yes", "yes")

films <- data.frame(srno, film, rate, lang_Eng)

str(films)

films <- films %>% 
  mutate(across(c(rate, lang_Eng), as.factor))

str(films)

```
# everthing 
select a key variable and everything

```{r}
library(gapminder)
gapminder %>% select(pop, everything())
```

# toupper
# tolower
```{r}
library(stringr)

data <- data.frame(Dose.Cm=c("d1", "D2", "D3"),
                Len.km=c("High", 'low', 'Low'))
glimpse(data)

data %>% mutate(Dose.Cm= tolower(Dose.Cm), Len.km=toupper(Len.km))

```

# factor

```{r}
data <- data.frame(Dose.Cm=c("d1", "D2", "D3"),
                Len.km=c("high", 'low', 'medium'))
data <- data %>% mutate(len= as.factor(Len.km))

glimpse(data)
```

# change order of factor

```{r}
data %>% mutate(len= fct_relevel(len, c('low', 'medium', 'high')))
```


# parse_number
This drops any non-numeric characters before or after the first number. The grouping mark specified by the locale is ignored inside the number.

```{r}

library(tidyverse)
class <- c('8th', '9th', '10th')
students <- c('25-30', '35-41', '21-28')
school <- data.frame(class, students)
school

glimpse(school) # notice students is a binned variable it is a not a numeric.

school %>% mutate(students= parse_number(students)) %>% glimpse()
school %>% mutate(students= parse_number(students))
# now students because number with first value of the column



```

# pivot longer
```{r}
library(tidyverse)

rawdata <- data.frame(species_1=rnorm(n = 40, mean = 300, sd = 18.5), species_2=rnorm(40, 305, 16.7))


data <- pivot_longer(data = rawdata, cols = species_1:species_2, names_to = 'species', values_to = 'weight')
```


# ggplot

```{r}
#sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
df <- data.frame(dose=c("D0.5", "D1", "D2"),
                len=c(4.2, 10, 29.5))
```
## bar plot 

```{r}
library(ggplot2)
# Basic barplot
p<-ggplot(data=df, aes(x=dose, y=len)) +
  geom_bar(stat="identity")
p
   
# Horizontal bar plot
# p + coord_flip()
```

```{r}
# Change the width of bars
ggplot(data=df, aes(x=dose, y=len)) +
  geom_bar(stat="identity", width=0.5)
# Change colors
ggplot(data=df, aes(x=dose, y=len)) +
  geom_bar(stat="identity", color="blue", fill="white")
# Minimal theme + blue fill color
p<-ggplot(data=df, aes(x=dose, y=len)) +
  geom_bar(stat="identity", fill="steelblue")+
  theme_minimal()
p
```

## geom_vline

```{r}
df <- data.frame(dose=c("D0.5", "D1", "D2", 'pp', 'kk', 'rr'),
                len=c(4.2, 10, 29.5, 12, 15, 23))
library(ggplot2)

ggplot(df, aes(len))+
  geom_density()+
  geom_vline(aes(xintercept = mean(len)), col='red', linetype= 'dashed')
```

## scatter plot with lm
```{r}
library(ggplot2)

ggplot(iris, aes(Petal.Length, Petal.Width))+
  geom_point()+
  geom_smooth(method = 'lm')
```
### raincloud plot
```{r}
library(ggdist)
library(tidyverse)
library(tidyquant)
mpg %>% filter(cyl %in% c(4,6,8)) %>% 
  ggplot(aes( x= factor(cyl), y= hwy, fill=factor(cyl)))+
# add half violin from `ggdist` package
ggdist::stat_halfeye(
  # custom bandwidth
  adjust = 0.5,
  # move geom to right
  justification= -0.2,
  # remove slab interval
  .width = 0,
  point_color= NA
)+
  # add boxplot
  geom_boxplot(
    width= 0.12,
    # remove outliers 
    outlier.colour = NA,
    alpha= 0.5
  )+
  # add dot plots from `ggdist` package
  ggdist::stat_dots(
    #orientation of the plot
    side= 'left',
    # move geom to the left
    justification= 1.1,
    # adjust grouping of observation
    binwidth=0.25
  )+
  # adjust theme
  scale_fill_tq()+
  theme_tq()+
  labs(
    title= 'raincloud plot',
    subtitle= 'showing bimodel distribution of 6 cylinder  vehicles',
    x= 'highway fuel efficiency',
    y= 'cylinders'
  )+
  coord_flip()
```


# months
```{r}
library(lubridate)
months <- seq(month(1:12)) # make moths
months <- month.abb[months] # make abbriviations
temperature <- c(10,12,22,32,35,30,33,28,29,25,19,14)
myframe <- data.frame(months,temperature) # creating a new data frame

library(tidyverse)
glimpse(myframe)

library(ggplot2)
ggplot(myframe, aes(x=months, y=temperature, group=1))+
  geom_line(col='blue')+
  geom_point(col='red')+
  ggtitle('Temperature of months')+ 
  scale_x_discrete(limits = month.abb) # this will order months on the x axis

```


```{r}
p + scale_x_discrete(limits=c("D0.5", "D2"))
```

```{r}
# Outside bars
ggplot(data=df, aes(x=dose, y=len)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=len), vjust=-0.3, size=3.5)+
  theme_minimal()
# Inside bars
ggplot(data=df, aes(x=dose, y=len)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=len), vjust=1.6, color="white", size=3.5)+
  theme_minimal()
```


```{r}
df2 <- data.frame(supp=rep(c("VC", "OJ"), each=3),
                dose=rep(c("D0.5", "D1", "D2"),2),
                len=c(6.8, 15, 33, 4.2, 10, 29.5))
```

```{r}
ggplot(data=df2, aes(x=dose, y=len, fill=supp)) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label=len), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()
```

```{r}
# Stacked barplot with multiple groups
ggplot(data=df2, aes(x=dose, y=len, fill=supp)) +
  geom_bar(stat="identity")
# Use position=position_dodge()
ggplot(data=df2, aes(x=dose, y=len, fill=supp)) +
geom_bar(stat="identity", position=position_dodge())
```

```{r}
# Change the colors manually
p <- ggplot(data=df2, aes(x=dose, y=len, fill=supp)) +
geom_bar(stat="identity", color="black", position=position_dodge())+
  theme_minimal()
# Use custom colors
p + scale_fill_manual(values=c('#999999','#E69F00'))
# Use brewer color palettes
p + scale_fill_brewer(palette="Blues")
```

# hex plot

```{r}
library(tidyverse)
# install.packages("hexbin")
class <- c(rep('10th', 8))
students <- c('10 to 15',
              "15-20",
              "17 to 24",
              "20  to 25",
              "25 to 30",
              "30 to 40",
              "45 to 47",
              '50 to 55')
latitude <- c(11.50897246,
              11.48323136,
              11.48719031,
              11.46366611,
              11.41097322,
              11.52111154,
              11.44491386,
              11.46569568)
longitude <- c(76.06032062,
               76.06192685,
               76.04266851,
               76.04156575,
               76.05075092,
               76.02846331,
               76.03084141,
               76.01766216)
school <- data.frame(class, students, latitude, longitude)

school %>% mutate(students= parse_number(students)) %>% 
  ggplot(aes(latitude, longitude, z= students))+
  stat_summary_hex()+
  scale_fill_viridis_c(alpha= 0.8)+
  labs(fill='students', title = 'school students')
```

# Color Palettes  

### Resources  

* https://colorhunt.co/
* https://coolors.co/
* https://colorpalettes.net/
* https://www.canva.com/colors/color-palettes/
* https://color.adobe.com/de/create/color-wheel
* https://mycolor.space/
* http://colormind.io/

```{r}
#install.packages(c("tidyverse", "gapminder", "MetBrewer"))
```


libraries
```{r}
library(tidyverse)
library(gapminder)
# install.packages('MetBrewer')
library(MetBrewer)

```

Plot the point plot using GDP per Capita as the x- axis and LE as the y axis. Numerical variable Population to control the size of each point.

```{r}
plot <- gapminder %>% 
  filter (year==2007) %>% 
  ggplot()+
  labs(x= 'GDP per Capita',
             y= 'Life Expectancy',
       color= 'Population in millions',
       size='Population in millions')+
  theme_minimal()

plot+ geom_point(aes(gdpPercap, lifeExp, size= pop/1000000))
```

To use color in the plot, assign the Population variable to the color aesthetic.
Since nothing is specied, ggplot2 chooses a color spectrum for this numerical variable (shades of blue).

```{r}
plot + geom_point(aes(gdpPercap, lifeExp, size= pop/1000000, color= pop/1000000))
```
To control the color spectrum, we need to introduce a color scale. In the
following plot, we have to provide a vector of hex color values. You would choose this if
you got your colors from one of the mentioned above websites.

```{r}
plot + geom_point(aes(gdpPercap, lifeExp, size= pop/1000000, color= pop/1000000))+
  scale_color_gradientn(colors = c("#003049", "#D62828", "#F77F00", "#FCBF49", "#EAE2B7"))
```

To apply one of the MetBrewer palettes, replace the hex-vector with a
MetBrewer function. Within the function call, you provide the palette’s name,
then several colors, and tell it that we need a continuous palette since it is a
numerical variable.

```{r}
plot + geom_point(aes(gdpPercap, lifeExp, size= pop/1000000, color= pop/1000000))+
  scale_color_gradientn(colors = met.brewer('Cross', n=500, type = 'continuous'))
```

You might also want to use color palettes with non-numerical variables. Let us
assume we want to apply color to the Continent variable. This implies using a
manual color scale and providing a MetBrewer palette.


```{r}
plot + geom_point(aes(gdpPercap, lifeExp, size= pop/1000000, color= continent))+
  scale_color_manual(values = met.brewer('Navajo', 5))
```

Please note if you want to apply color to the fill aesthetic rather than the color
aesthetic, consider using the scale_fill_manuel function instead of the
scale_color_manuel. This is useful for boxplots or bar charts. 

```{r}
box <- gapminder %>% 
  filter(gdpPercap< 60000) %>% 
  ggplot(aes(continent, gdpPercap, color= year, fill= continent))+
           geom_boxplot()+
  theme_minimal()+ labs( x= 'Continent', y= 'GDP per Capita', fill= 'Continent')
  
```

# scale fill manual
# themes
```{r}
df <- data.frame(
    Names=as.factor(c('Bacteria', 'Yeast', 'None')),
    Quantity=c(2.5, 5.5, 7.5))

library(ggplot2)
library(tidyverse)
df <- df %>% mutate(Names= fct_relevel(Names, c('Bacteria', 'Yeast', 'None')))

ggplot(df, aes(Names, Quantity, fill= Names))+
    geom_bar(stat = 'identity')+
    scale_fill_manual(values = c('#110a62', '#fcd749','#b5b4b5'))+
  labs(y='Necter pH', x= 'Microbe added to nectar')+
  theme_classic()+
    theme(legend.position = 'none', axis.ticks.x = element_blank())+
  theme(axis.text = element_text(size = 22, color= 'black'))+
theme(axis.line.x = element_blank())+
  theme(axis.ticks = element_line(size = 1, color="black"), 
   axis.ticks.length = unit(.5, "cm"))+
  theme(text = element_text(size = 22))

```


### graphics
```{r}
x11() # opne a new window for graphics
graphics.off() # close the new window 
```

# Normal distribution  

Normal distribution, also known as the Gaussian distribution, is a probability distribution that is symmetric about the mean, showing that data near the mean are more frequent in occurrence than data far from the mean.

```{r}
library(tidyverse)
n = 1000
mean = 170 # cm
sd = 6.35 # cm
binwidth= 0.3
set.seed(1234)
df <- data.frame(x=rnorm(n, mean, sd))
ggplot(df, aes(x = x, mean = mean, sd = sd, binwidth = binwidth, n = n))+
    theme_bw()+
  geom_histogram(binwidth = binwidth, 
        colour = "white", fill = "lightblue", size = 0.1)+
stat_function(fun = function(x) dnorm(x, mean = mean, sd = sd) * n * binwidth,
    color = "darkred", size = 1)

```

# Functions

### dice
```{r}
dice <- c(1:6)

myluck<- function(x){
  myluck <- sample(dice, size = 1, replace = T)
  return(myluck)
}

myluck()
```
### pick a name

```{r}
names <- c('saneesh','appu','sanusha')
who <- function(x){
  who <- sample(names, 1, T)
  return(who)
}

who()
```

# function to split
```{r}

df <- data.frame(name=as.factor(c('James Bond', 'Spider Man', 'Iron Man')))
# df <- df %>% separate(name, c('Genus', 'Species'), sep = '([ ])')

shorten <- function(df){
  name_split <- df %>% separate(name, c('Genus', 'Species'), sep = '([ ])')
  print(name_split)
}

shorten(df)
```

# Rmarkdown  
### knitr golbal options
to apply to every chunk in the file

inside the `chunk` write `knitr::opts_chunk$set(include= ,echo = , message= , warning= )`

```{r}
# knitr::opts_chunk$set(message = TRUE, echo = TRUE, warning = TRUE)
```

`include:` to show or hide code and results from appearing        
`echo:` to show or hide code in the output but shows result       
`message` to hide or show the messages generated by the code        
`warning:` to show or hide warning generated by the code  

these options can be written for individual chunks as well

```{r echo=FALSE}
sum(2,3) # here the knitr global option is over written inside the {}
```


# Heading 1  
## Heading 2
### Heading 3


*italics*   
_italic_  

**bold**  
__bold__  

`plot()` to show r code/function  
@Saneesh

## blockquotes are writtedn after >

> this is a blockquote  
>--- Saneesh

### plain code

```
hello
```

### unordered items  

- item 1  
- item 2  
  - sub item 1a  
  - sub item 2b  

### ordered items
1. Item 1  
2. Item 2  
   + Item 2a  # give two spaces before the +
   + Item 2b  

## writing mathematical functions

### adding a link
[mathematical notations](https://rpruim.github.io/s341/S19/from-class/MathinRmd.html)

$by$  
$\mu$  
$\sum$   
$a\pm b$    
$x=y$  
$x>y$  
$x^2$  
$x\le y$  
$\sum_{n=1}^{10} n^2$  
$x_{1}+ x_{2}+\cdots+x_{n}$  
$|A|$  
$A\subset B$  
$A \subseteq B$  
$A \cup B$  
$A \cap B$  
$P(A|B)$  
$\alpha$  
$\beta$  
$\gamma$  
$\theta$  
$H_2O$
 

### table
using `knitr::kable()`

```{r table, echo=FALSE}
library(dplyr)
iris %>% filter(Species %in% ('virginica'), Sepal.Length== 6.3) %>% 
  knitr::kable()

```




